Model Checking Differentially Private Properties

### 2018-01-30 examples.tex
work through and 16 pages ;)
adaptions

\begin{example}\label{exa:survey}
  Figure~\ref{figure:2-dp} shows the survey
mechanism and its corresponding Markov chain. In the figure, the states
$+$ and $-$ denote the diagnoses are positive or negative
respectively; the states $s$ and $t$ denote answers to the survey
question and hence let $\mathit{out}_Y \in L (s)$ and $\mathit{out}_N
\in L (t)$.
For each state $x$, we moreover associate an atomic predicate
$\mathit{at}_x$ (not shown in the figure). 
States $+$ and $-$ are neighbors. 
The random noise boosts the probability of answering
\textit{Yes} or \textit{No} to at least $1/4$ regardless of
diagnoses. Inferences on individual diagnosis can be plausibly denied.
In general, missing transitions (such as those from $Y$ and $N$) lead
to a special state $\dagger$ with a self-loop. We omit the transitions
and the state $\dagger$ for clarity.
\end{example}

@@

\begin{algorithm}
	\begin{algorithmic}[1]
		\Function{OnlineModel}{$d$, $t$}
		\Match{the input threshold $t$}
		\Comment{Assignments of $t'$ by $\frac{1}{4}$-geometric mechanism}
		\lCase{$0$}{$t' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{4}{5}$,$\frac{3}{20}$,
			$\frac{3}{80}$,$\frac{3}{320}$,
			$\frac{3}{1280}$,$\frac{1}{1280}$ respectively}
		\lCase{$1$}{$t' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{1}{5}$,$\frac{3}{5}$,
			$\frac{3}{20}$,$\frac{3}{80}$,
			$\frac{3}{320}$,$\frac{1}{320}$ respectively}
		\lCase{$2$}{$t' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{1}{20}$,$\frac{3}{20}$,
			$\frac{3}{5}$,$\frac{3}{20}$,
			$\frac{3}{80}$,$\frac{1}{80}$ respectively}
		\lCase{$3$}{$t' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{1}{80}$,$\frac{3}{80}$,
			$\frac{3}{20}$,$\frac{3}{5}$,
			$\frac{3}{20}$,$\frac{1}{20}$ respectively}
		\lCase{$4$}{$t' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{1}{320}$,$\frac{3}{320}$,
			$\frac{3}{80}$,$\frac{3}{20}$,
			$\frac{3}{5}$,$\frac{1}{5}$ respectively}
		\lCase{$5$}{$t' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{1}{1280}$,$\frac{3}{1280}$,
			$\frac{3}{320}$,$\frac{3}{80}$,
			$\frac{3}{20}$,$\frac{4}{5}$ respectively}
		\EndMatch
		\For{each query $f_i$}
		\Match{the input databese $d$}
		\Comment{Assignments of $f_{i}'$ by $\frac{1}{2}$-geometric mechanism}
		\lCase{$0$}{$f_{i}' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{2}{3}$,$\frac{1}{6}$,
			$\frac{1}{12}$,$\frac{1}{24}$,
			$\frac{1}{48}$,$\frac{1}{48}$ respectively}
		\lCase{$1$}{$f_{i}' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{1}{3}$,$\frac{1}{3}$,
			$\frac{1}{6}$,$\frac{1}{12}$,
			$\frac{1}{24}$,$\frac{1}{24}$ respectively}
		\lCase{$2$}{$f_{i}' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{1}{6}$,$\frac{1}{6}$,
			$\frac{1}{3}$,$\frac{1}{6}$,
			$\frac{1}{12}$,$\frac{1}{12}$ respectively}
		\lCase{$3$}{$f_{i}' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{1}{12}$,$\frac{1}{12}$,
			$\frac{1}{6}$,$\frac{1}{3}$,
			$\frac{1}{6}$,$\frac{1}{6}$ respectively}
		\lCase{$4$}{$f_{i}' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{1}{24}$,$\frac{1}{24}$,
			$\frac{1}{12}$,$\frac{1}{6}$,
			$\frac{1}{3}$,$\frac{1}{3}$ respectively}
		\lCase{$5$}{$f_{i}' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{1}{48}$,$\frac{1}{48}$,
			$\frac{1}{24}$,$\frac{1}{12}$,
			$\frac{1}{6}$,$\frac{2}{3}$ respectively}
		\EndMatch
		\If{$f_{i}'>t'$}
		\State{$a_i=\top$}
		\State{\Return}
		\Else
		\State{$a_i=\bot$}
		\EndIf
		\EndFor
		\EndFunction

	\end{algorithmic}
	\caption{Input is a private database $d \in \{0,1,2,3,4,5\}$, an adaptively chosen stream of sensitivity 1 queries $f_1$, . . ., a threshold $t \in \{0,1,2,3,4,5\}$.Output is a stream of answers $a_1$,...}
	\label{algorithm:online-model}
\end{algorithm}

@@

\paragraph{Online Model.}
Below we describe an online model adapted from~\cite{DR:14:AFDP}: it is used in the scenario where a large amount of queries occur, but we only care for the queries which
are above a given threshold. Other queries below the threshold are considered "irrelevent" and might be discarded. In ~\cite{DR:14:AFDP}, since the model is set for continuous database, the Laplace mechanism is applied to both queries and the threshold. In our model, the database is discrete and instead the geometric mechanism is applied.

We consider the database set $D = \{0,1,2,3,4,5\}$, a threshold $t \in \{0,1,2,3,4,5\}$, queries $\{f| f $ with sensitivity $\Delta (f) = 1 \}$. In order to protect privacy, truncated $\alpha$-geometric mechanisms over $\underline{5}$ is applied to each query and the threshold, and what we observe is the first time that a query result is greater than or equal to the threshold after interference. We are interested in the differential privacy property of the data in $D$, assuming that a pair of data $d,d'$are neighbours, if $|d-d'|\le 1$. The algorithm is shown in Algorithm~\ref{algorithm:online-model}.

>>>>

\begin{example}\label{exa:survey}
  Figure~\ref{figure:2-dp} shows the survey
mechanism and its corresponding Markov chain. In the figure, the states
$+$ and $-$ denote the diagnoses are positive or negative
respectively; the states $s$ and $t$ denote answers to the survey
question and hence let $\mathit{out}_Y \in L (s)$ and $\mathit{out}_N
\in L (t)$.
States $+$ and $-$ are neighbors.
The random noise boosts the probability of answering
\textit{Yes} or \textit{No} to at least $1/4$ regardless of
diagnoses. Inferences on individual diagnosis can be plausibly denied.
In general, missing transitions (such as those from $Y$ and $N$) lead
to a special state $\dagger$ with a self-loop. We omit the transitions
and the state $\dagger$ for clarity.
\end{example}

@@

\begin{algorithm}[tbh]
	\begin{algorithmic}[1]
		\Function{OnlineModel}{$d$, $t$}
		\Match{the input threshold $t$}
		\Comment{Assignments of $t'$ by $\frac{1}{4}$-geometric mechanism}
		\lCase{$0$}{$t' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{4}{5}$,$\frac{3}{20}$,
			$\frac{3}{80}$,$\frac{3}{320}$,
			$\frac{3}{1280}$,$\frac{1}{1280}$ respectively}
		\lCase{$1$}{$t' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{1}{5}$,$\frac{3}{5}$,
			$\frac{3}{20}$,$\frac{3}{80}$,
			$\frac{3}{320}$,$\frac{1}{320}$ respectively}
		\lCase{$2$}{$t' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{1}{20}$,$\frac{3}{20}$,
			$\frac{3}{5}$,$\frac{3}{20}$,
			$\frac{3}{80}$,$\frac{1}{80}$ respectively}
		\lCase{$3$}{$t' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{1}{80}$,$\frac{3}{80}$,
			$\frac{3}{20}$,$\frac{3}{5}$,
			$\frac{3}{20}$,$\frac{1}{20}$ respectively}
		\lCase{$4$}{$t' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{1}{320}$,$\frac{3}{320}$,
			$\frac{3}{80}$,$\frac{3}{20}$,
			$\frac{3}{5}$,$\frac{1}{5}$ respectively}
		\lCase{$5$}{$t' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{1}{1280}$,$\frac{3}{1280}$,
			$\frac{3}{320}$,$\frac{3}{80}$,
			$\frac{3}{20}$,$\frac{4}{5}$ respectively}
		\EndMatch
		\For{each query $f_i$}
		\Match{the input databese $d$}
		\Comment{Assignments of $f_{i}'$ by $\frac{1}{2}$-geometric mechanism}
		\lCase{$0$}{$f_{i}' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{2}{3}$,$\frac{1}{6}$,
			$\frac{1}{12}$,$\frac{1}{24}$,
			$\frac{1}{48}$,$\frac{1}{48}$ respectively}
		\lCase{$1$}{$f_{i}' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{1}{3}$,$\frac{1}{3}$,
			$\frac{1}{6}$,$\frac{1}{12}$,
			$\frac{1}{24}$,$\frac{1}{24}$ respectively}
		\lCase{$2$}{$f_{i}' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{1}{6}$,$\frac{1}{6}$,
			$\frac{1}{3}$,$\frac{1}{6}$,
			$\frac{1}{12}$,$\frac{1}{12}$ respectively}
		\lCase{$3$}{$f_{i}' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{1}{12}$,$\frac{1}{12}$,
			$\frac{1}{6}$,$\frac{1}{3}$,
			$\frac{1}{6}$,$\frac{1}{6}$ respectively}
		\lCase{$4$}{$f_{i}' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{1}{24}$,$\frac{1}{24}$,
			$\frac{1}{12}$,$\frac{1}{6}$,
			$\frac{1}{3}$,$\frac{1}{3}$ respectively}
		\lCase{$5$}{$f_{i}' \leftarrow 0, 1,2,3,4,5$ with probability
			$\frac{1}{48}$,$\frac{1}{48}$,
			$\frac{1}{24}$,$\frac{1}{12}$,
			$\frac{1}{6}$,$\frac{2}{3}$ respectively}
		\EndMatch
		\If{$f_{i}'>t'$}
		\State{$a_i=\top$}
		\State{\Return}
		\Else
		\State{$a_i=\bot$}
		\EndIf
		\EndFor
		\EndFunction

	\end{algorithmic}
	\caption{Input is a private database $d \in \{0,1,\ldots,5\}$, an adaptively chosen stream of sensitivity 1 queries $f_1,f_2$,\ldots, a threshold $t \in \{0,1,\ldots,5\}$.Output is a stream of answers $a_1,a_2$,\ldots}
	\label{algorithm:online-model}
\end{algorithm}

@@

\paragraph{Online Model.}
Below we describe an online model adapted from~\cite{DR:14:AFDP}: it is used in the scenario where a large amount of queries occur, but we only care for the queries which
are above a given threshold. Other queries below the threshold are considered \emph{irrelevant} and might be discarded. In~\cite{DR:14:AFDP}, the model is set for continuous database, and the Laplace mechanism is applied to both queries and the threshold. In our model, the database is discrete and instead the geometric mechanism is applied.

We consider the database set $D = \{0,1\ldots,5\}$, a threshold $t \in \{0,1\ldots,5\}$, and queries $\{f| f $ with sensitivity $\Delta (f) = 1 \}$. In order to protect privacy, truncated $\alpha$-geometric mechanisms over $\{0,1\ldots,5\}$ is applied to each query and the threshold, and what we observe is the first time that a query result is greater than or equal to the threshold after interference. We are interested in the differential privacy property of the data in $D$, assuming that a pair of data $d,d'$are neighbours, if $|d-d'|\le 1$. The algorithm is shown in Algorithm~\ref{algorithm:online-model}.

### 2018-01-30 semantics.tex
adaptions

Given a finite discrete-time Markov chain $K = (S, \wp, L)$, a
\emph{neighborhood relation on $S$} $\neighbor{S} \subseteq S \times S$
is a reflexive and symmetric relation on $S$. We will write $s
\neighbor{S} t$ when $(s, t) \in \neighbor{S}$. If $s \neighbor{S} t$,
we say $s$ and $t$ are \emph{neighbors} or $t$ is a \emph{neighbor} of
$s$. Define the satisfaction relation $K, \neighbor{S}, s
\models \Phi$ as follows.
\begin{eqnarray*}
%  K, \neighbor{S}, s \models \top\\
  K, \neighbor{S}, s \models p
  & \textmd{ if } &
  p \in L(s)\\
  K, \neighbor{S}, s \models \neg \Phi
  & \textmd{ if } &
  K, \neighbor{S}, s \not\models \Phi\\
  K, \neighbor{S}, s \models \Phi_0 \wedge \Phi_1
  & \textmd{ if } &
  K, \neighbor{S}, s \models \Phi_0 \textmd{ and }
  K, \neighbor{S}, s \models \Phi_1\\
  K, \neighbor{S}, s \models \PJ{J} \phi
  & \textmd{ if } &
  \Pr [\{ \pi : K, \neighbor{S}, \pi \models \phi \textmd{ with }
                    \pi_0 = s\}] \in J \\
  K, \neighbor{S}, s \models \dpriv{\epsilon}{\delta} \phi
  & \textmd{ if } &
  \textmd{for every } t \textmd{ with }  s \neighbor{S} t,
      p(s) \leq e^{\epsilon} p(t) + \delta \textmd{ and }
      p(t) \leq e^{\epsilon} p(s)  +  \\
  & &  \delta \textmd{ where } p(x) = \Pr [\{
      \pi : K, \neighbor{S}, \pi \models \phi \textmd{ with }
      \pi_0 = x \}]
\end{eqnarray*}

@@

In the classical $\PCTL$ model checking algorithm for Markov chains,
states satisfying the subformula $\PJ{J} \phi$ are obtained by
computing $\myPr{s}{K}{\neighbor{S}}{\phi}$ for state $s \in S$.
These probabilities can be computed by solving linear equations or
through iterative approximations. We summarize it in the following
theorem~\cite{BK:08:PMC}:

\begin{lemma}
  Let $K = (S, \wp, L)$ be a Markov chain, $s
  \in S$, and $B, C \subseteq S$. The probabilities
  $\myPr{s}{K}{\neighbor{S}}{\gX B}$ and
  $\myPr{s}{K}{\neighbor{S}}{B \mho C}$
\hide{
  $\Pr[\{ \pi : K, \neighbor{K}, \pi \models B \buntil{n} C \textmd{
    with } \pi_0 = s \}]$
} are computable within time polynomial in
  $|S|$.
  \label{lemma:PJ-subroutines}
\end{lemma}

@@

The model checking algorithm for $\dpCTLstar$ can be treated as in the classical setting~\cite{BK:08:PMC}:  all we need is to  compute the  probability
  $\myPr{s}{K}{\neighbor{S}}{\phi}$ with general path formula $\phi$. For this purpose one first constructs a deterministic $\omega$-automaton $R$, such as deterministic Rabin automaton. Then, the probability reduces to a
  reachability probability in the product Markov chain obtained from $K$ and $R$. There are more efficient algorithms without the product construction, see \cite{CY95,CSS03,BaierKKKMW16} for details.

>>>>

Given a finite Markov chain $K = (S, \wp, L)$, a
\emph{neighborhood relation} $\neighbor{S} \subseteq S \times S$
is a reflexive and symmetric relation on $S$. We will write $s
\neighbor{S} t$ when $(s, t) \in \neighbor{S}$. If $s \neighbor{S} t$,
we say $s$ and $t$ are \emph{neighbors} or $t$ is a \emph{neighbor} of
$s$. Define the satisfaction relation $K, \neighbor{S}, s
\models \Phi$ as follows.
\begin{eqnarray*}
%  K, \neighbor{S}, s \models \top\\
  K, \neighbor{S}, s \models p
  & \textmd{ if } &
  p \in L(s)\\
  K, \neighbor{S}, s \models \neg \Phi
  & \textmd{ if } &
  K, \neighbor{S}, s \not\models \Phi\\
  K, \neighbor{S}, s \models \Phi_0 \wedge \Phi_1
  & \textmd{ if } &
  K, \neighbor{S}, s \models \Phi_0 \textmd{ and }
  K, \neighbor{S}, s \models \Phi_1\\
  K, \neighbor{S}, s \models \PJ{J} \phi
  & \textmd{ if } &
  \Pr [\{ \pi : K, \neighbor{S}, \pi \models \phi \textmd{ with }
                    \pi_0 = s\}] \in J \\
  K, \neighbor{S}, s \models \dpriv{\epsilon}{\delta} \phi
  & \textmd{ if } &
  \textmd{for every } t \textmd{ with }  s \neighbor{S} t,
      p(s) \leq e^{\epsilon} p(t) + \delta \textmd{ and }
      p(t) \leq e^{\epsilon} p(s)  +  \\
  & &  \delta \textmd{ where } p(x) = \Pr [\{
      \pi : K, \neighbor{S}, \pi \models \phi \textmd{ with }
      \pi_0 = x \}]
\end{eqnarray*}

@@

In the classical $\PCTL$ model checking algorithm for Markov chains,
states satisfying the subformula $\PJ{J} \phi$ are obtained by
computing $\myPr{s}{K}{\neighbor{S}}{\phi}$ for state $s \in S$.
These probabilities can be computed by solving linear equations or
through iterative approximations. We summarize it in the following
theorem (details see~\cite{BK:08:PMC}):

\begin{lemma}
  Let $K = (S, \wp, L)$ be a Markov chain, $s
  \in S$, and $B, C \subseteq S$. The probabilities
  $\myPr{s}{K}{\neighbor{S}}{\gX B}$ and
  $\myPr{s}{K}{\neighbor{S}}{B \mho C}$ 
\hide{
  $\Pr[\{ \pi : K, \neighbor{K}, \pi \models B \buntil{n} C \textmd{
    with } \pi_0 = s \}]$
} are computable within time polynomial in
  $|S|$.\lz{and $|\wp|$?}
  \label{lemma:PJ-subroutines}
\end{lemma}

@@

The model checking algorithm for $\dpCTLstar$ can be treated as in the classical setting~\cite{BK:08:PMC}:  all we need is to  compute the  probability
  $\myPr{s}{K}{\neighbor{S}}{\phi}$ with general path formula $\phi$. For this purpose one first constructs a deterministic $\omega$-automaton $R$ corresponding to $\phi$, such as deterministic Rabin automaton. Then, the probability reduces to a
  reachability probability in the product Markov chain obtained from $K$ and $R$. There are more efficient algorithms without the product construction, see \cite{CY95,CSS03,BaierKKKMW16} for details.

### 2018-01-30 specifications.tex
adaptions

As an example, for our threshold \lz{to do} model, we consider the property $\wedge_{i=1}^{7} \dpriv{3}{0}(\X^i \top) \wedge \PJ{\le .1}(\X^7 F T) \wedge \dpriv{3}{0}(\X^7 F \top)$.
that is used to express the privacy for our threshold Markov chains. It is only a valid formula if we would allow infinite conjunctions. However, it is interesting to note the relation to language equivalence for probabilistic automata\lz{to be discussed}

>>>>

As an example, for our threshold model (Algorithm~\ref{algorithm:online-model}), we consider the property $\Phi=\wedge_{i=1}^{6} \dpriv{3}{0}(\X^i \top) \wedge \PJ{\le 0.025}(\X^6 F \top) \wedge \dpriv{3}{0.025}(\X^6 F \top)$, to express the privacy. The state $s_{34}$ (with the input threshold $t=3$, and the database $d=4$) satisfies $\Phi$ by our computation. It means the probability of reaching $\top$  within $i$ steps with $i$ ranges over $\{1,\ldots,6\}$, is $(3, 0)$-differentially private for the database $d$ and its neighbours ($d=3$ and $d=5$). It also states that the probability of reaching $\top$ after no less than $6$ steps, is smaller than $0.025$ and is  $(3, 0.025)$-differentially private.
